Subsidiary Recursion in Coq

### 2022-01-30 paper/paper.tex
bug fixes, small improvements
Almost in shape to submit to ITP 2022

Algebras for our subsidiary recursion are more complex than
$F$-algebras.  For reasons we will explain further below, the carrier
of the algebra will be a functor \verb|X : Set -> Set|.  The algebra
is presented with:

@@

Our development is parametrized by a signature functor \verb|F| of
type \verb|Set -> Set|.  It implements \verb|Subrec : Set|, which is
the type to use for subsidiary recursion.  This type comes with
\verb|inn : F Subrec -> Subrec|, which behaves computationally like a
constructor.  \verb|Subrec.v| does not itself provide an induction
principle based on \verb|inn|, however.  Induction is derived later
(Section~\ref{sec:interfacei}). \verb|Subrec.v| makes critical use of
retractive-positive recursive types, to take a fixed-point of a
construction based on \verb|F|.  We present these recursive types in
Section~\ref{sec:mu} below.

For our examples, we will consider the specialization to the case of
lists, parametrized by the type \verb|A| of elements.  In general, to
use our development to get subsidiary recursion over some datatype,
one must define a signature functor for the datatype.  For lists, this
is \verb|ListF| of Figure~\ref{fig:listf}.  \verb|List| is defined to
be \verb|Subrec|, with \verb|F| instantiated to \verb|ListF A|.  This
type \verb|List| is not to be confused with the type \verb|list| of
lists in Coq's standard library.  As noted previously, our development
is meant to be used in extension of existing inductive datatypes, not
replacing them.  The figure also shows constructors \verb|mkNil| and
\verb|mkCons| for \verb|List|, and types for conversion functions
between \verb|List| and \verb|list| (code elided). One direction uses
Coq's structural recursion, the other uses subsidiary recursion, which
we will see next.

@@

Finally, for a recursion scheme, one would like to see not just the
typed interface, but also the computation law.  This is stated as a
theorem in Figure~\ref{fig:comp}.  Intuitively, it states that
\verb|fold|ing an algebra over constructed data \verb|inn d| is equal
to invoking the algebra on \verb|fold|
for the fold function; an invocation of \verb|fold| with the algebra
for the \verb|rec| function; and \verb|d| for the subdata structure.

\begin{figure}
\begin{minted}{coq}
Theorem FoldChar :
 forall (X : Set -> Set) (FunX : Functor X) (IdF : FmapId X FunX)
        (algf : AlgF Alg X) (d : F Subrec),
 fold X FunX (rollAlg algf) (inn d) =
   algf _ fold (fold X FunX (rollAlg algf)) d .
\end{minted}
\end{figure}

@@

\begin{figure}
\begin{minted}{coq}
  Definition WordsBy(p : A -> bool)
    : Alg (ListF A) (Const (list (list A))) :=
    rollAlg (fun R fo wordsBy xs => 
      match xs with
        Nil => [] 
      | Cons hd tl =>
        if p hd then
          wordsBy tl
        else
          let (w,z) := breakr fo p tl in
            (hd :: w) :: wordsBy z
      end).

Definition wordsByr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                   (p : A -> bool)(xs : R) : list (list A) :=
  fo (Const (list (list A))) (FunConst (list (list A))) (WordsBy p) xs.
\end{minted}
\caption{Functions  \texttt{wordsBy} and \texttt{wordsByr}, and the algebra they
  fold (\texttt{WordsBy.v})}
\label{fig:wordsby}
\end{figure}

@@

\begin{figure}
\begin{minted}{coq}
Definition MapThroughAlg{B : Set}(f:mappedT A B) : Alg (ListF A) (Const (list B)) :=
  rollAlg (fun R fo mapThrough xs => 
    match xs with
      Nil => []
    | Cons hd tl =>
      let (b,c) := f R fo hd tl in
        b :: mapThrough c
    end).

Definition mapThroughr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                      {B : Set}(f:mappedT A B) : R -> list B.
Definition mapThrough{B : Set}(f:mappedT A B) : List A -> list B.
\end{minted}
\caption{The algebra \texttt{MapThroughAlg} defining function \texttt{mapThrough} and \texttt{mapThroughr}; the
  code for those follows the pattern of \texttt{wordsBy} and \texttt{wordsByr} (Figure~\ref{fig:wordsby}), so
  we omit it (\texttt{MapThrough.v})}
\label{fig:mapthrough}
\end{figure}

@@

Assuming \verb|A : Set| and an equality test \verb|eqb : A -> A -> bool| on it,
we port this code to our Coq infrastructure in Figure~\ref{fig:rle}.
The function \verb|compressSpan| is written at the type \verb|mappedT A (nat * A)| that
will be required by \verb|mapThrough|.  Unfolding the definition of \verb|mappedT|,
\verb|compressSpan| has type:
\begin{minted}{coq}
forall(R : Set)(fo:FoldT (Alg (ListF A)) R), A -> R -> (nat * A) * R.
\end{minted}
\noindent It will be invoked by the code for \verb|mapThrough| with a
fold function \verb|fo| of type \verb|FoldT (Alg (ListF A)) R|.  It then has the
responsibility of extracting from the tail at type \verb|R| (second input) 
a result upon which \verb|mapThrough| should recurse (second component
of the output pair).  Then we define an algebra \verb|RleAlg| by
supplying \verb|compressSpan| as the function to map through, to
\verb|MapThroughAlg| (Figure~\ref{fig:mapthrough}).  Following the
pattern seen above, we define function \verb|rle| for top-level
recursions using \verb|fold| (we could also define a subsidiary version
\verb|rler|).

@@

The definition of \verb|Mu| weakens
this to a strictly positive approximation:
\begin{minted}{coq}
  Inductive Mu : Set := 
    mu : forall (R : Set), (R -> Mu) -> F R -> Mu.
\end{minted}
\noindent Instead of taking in \verb|F Mu|, constructor \verb|mu|
accepts an input of type \verb|F R|, for some type \verb|R| for which
we have a function of type \verb|R -> Mu|.  The impredicative
quantification of \verb|R| is essential here: we instantiate it with
\verb|Mu| itself in the definition of \verb|inMu|
(Figure~\ref{fig:mu}).  So this approach would not work in a
predicative theory like Agda's.  The quantification of \verb|R| can be
seen as applying a technique due to Mendler, of introducing
universally quantified variables for problematic type occurrences, to
a datatype constructor.  We will review this in
Section~\ref{sec:related}.

@@

For a simple example: suppose we instantiate \verb|F| with
\verb|ListF| (of Figure~\ref{fig:listf}).  Our derivation uses a
different type that wraps \verb|F|, but this will show the issue in a
simple form.  Let us temporarily define \verb|List A| as
\verb|Mu (ListF A)| (again, for subsidiary recursion do not use just
\verb|ListF| directly).  The canonical way to define the empty list
would be, implicitly instantiating \verb|F| to \verb|ListF A|,
\begin{minted}{coq}
Definition mkNil := mu (List A) (fun x => x) (NilF A)
\end{minted}
\noindent But given this, there are infinitely many other equivalent
definitions.  For any \verb|Q : Set|, we could take
\begin{minted}{coq}
Definition mkNil' := mu Q (fun x => mkNil) (NilF A)
\end{minted}
\noindent Since \verb|fmap f (NilF A)| equals just \verb|NilF B| for
\verb|f : A -> B|, if we apply \verb|outMu| (of Figure~\ref{fig:mu})
to \verb|mkNil'| or \verb|mkNil|, we will get \verb|NilF (List A)|.
But critically, \verb|mkNil| and \verb|mkNil'| are not equal, neither
definitionally nor provably.  One can define a function that puts
\verb|Mu| values in normal form by folding \verb|inMu| over them.
Then \verb|mkNil| and \verb|mkNil'| will have the same normal form,
and be equivalent in that sense.  But the fact that they are not
provably equal is what we term noncanonicity.

@@

Our work contributes to the program proposed by Owens and Slind, of
broadening the scope of functional programs that can be accommodated
in ITPs~\cite{owens+08}.  The goal of terminating recursion has been
advocated in the literature on programming languages under the name
\emph{strong functional programming}~\cite{turner95}.  Uustalu and
Vene developed a categorical view of a recursion scheme allowing one
level of subsidiary recursion, and illustrated it in Haskell with an
artificial example~\cite{uustalu11}.  In contrast, our scheme allows
arbitrary finite nestings of recursion, and we illustrate it in Coq
with realistic examples.  It seems that generalizing the carriers of
algebras to functors is the critical step enabling such examples.
Like Uustalu and Vene, we find that subsidiary recursion subsumes
course-of-value recursion.  Our method is also similar to the
technique of sized types, in providing a type-based method for
termination~\cite{barthe04}.  With sized types, datatypes are indexed
with abstract sizes, which must then be propagated through code, using
dependent types.  Our \verb|Subrec| does not require dependent types,
resulting in just polymorphically typed code for our examples
(\verb|Subreci|, for proofs, of course does use dependent types).

\textbf{Mendler-style recursion.}  Mendler introduced the idea of
using universal abstraction to support compositional termination
checking~\cite{mendler91}.  He proposed a functor-generic recursor of
type $\all{X}{(\all{R}{(R \to X) \to F\ R \to X}) \to \mu\ F\ \to
  X}$. We have adopted this idea to the constructor of the type
\verb|Mu| (Section~\ref{sec:mu}).  Previous work explored the
categorical perspective on Mendler-style recursion, and showed how to
reduce it to basic catamorphisms (i.e., structural
recursion)~\cite{uustalu99}.  Another considered its use with negative
type schemes~\cite{ahn11}.  Previous work from our group showed how to
derive inductive datatypes in Cedille using extensions of the Mendler
encoding~\cite{firsov+18b,firsov+18a}.  Here, we do not derive
inductive types using these methods, but rather apply them to justify
a terminating recursion scheme for existing datatypes.

>>>>

Algebras for our subsidiary recursion are more complex than
$F$-algebras.  Let us begin with an informal explanation.  For reasons
we will explain further below, the carrier of the algebra will be a
functor \verb|X : Set -> Set|.  The algebra is presented with:

@@

The type over which one can recurse with our scheme of subsidiary
recursion is called \verb|Subrec|.  It is parametrized by a signature
functor \verb|F| of type \verb|Set -> Set|.  \verb|Subrec| comes with
\verb|inn : F Subrec -> Subrec|, which behaves computationally like a
constructor.  We will later derive an induction principle for this type
(Section~\ref{sec:interfacei}). The definition of \verb|Subrec| uses
retractive-positive recursive types, to take a fixed-point of a
construction based on \verb|F|.  We present these recursive types in
Section~\ref{sec:mu} below.

For our examples, we will consider the specialization to the case of
lists, parametrized by the type \verb|A| of elements.  In general, to
use our development to get subsidiary recursion over some datatype,
one must define a signature functor for the datatype.  For lists, this
is \verb|ListF|, which we saw at the start of Section~\ref{sec:interface}.  \verb|List| is defined to
be \verb|Subrec|, with \verb|F| instantiated to \verb|ListF A|.  This
type \verb|List| is not to be confused with the type \verb|list| of
lists in Coq's standard library.  As noted previously, our development
is meant to be used in extension of existing inductive datatypes, not
replacing them.  The figure also shows constructors \verb|mkNil| and
\verb|mkCons| for \verb|List|, and typings for conversion functions
between \verb|List| and \verb|list| (code elided).

@@

Finally, for a recursion scheme, one would like to see not just the
typed interface, but also the computation law.  This is shown as a
theorem in Figure~\ref{fig:comp}.  Intuitively, it states that
\verb|fold|ing an algebra over constructed data \verb|inn d| is equal
to invoking the algebra on \verb|fold|
for the fold function; an invocation of \verb|fold| with the algebra
for the \verb|rec| function; and \verb|d| for the subdata structure.

\begin{figure}
\begin{minted}{coq}
Theorem FoldChar :
 forall (X : Set -> Set) (FunX : Functor X) (IdF : FmapId X FunX)
        (algf : AlgF Alg X) (d : F Subrec),
 fold X FunX (rollAlg algf) (inn d) =
   algf _ fold (fold X FunX (rollAlg algf)) d .
\end{minted}
\caption{Computation law for subsidiary recursion, stated as a theorem}
\label{fig:comp}
\end{figure}

@@

\begin{figure}
\begin{minted}{coq}
  Definition WordsBy(p : A -> bool)
    : Alg (ListF A) (Const (list (list A))) :=
    rollAlg (fun R fo wordsBy xs => 
      match xs with
        Nil => [] 
      | Cons hd tl =>
        if p hd then
          wordsBy tl
        else
          let (w,z) := breakr fo p tl in
            (hd :: w) :: wordsBy z
      end).
Definition wordsByr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                   (p : A -> bool)(xs : R) : list (list A) :=
  fo (Const (list (list A))) (FunConst (list (list A))) (WordsBy p) xs.
\end{minted}
\caption{Functions  \texttt{wordsBy} and \texttt{wordsByr}, and the algebra they
  fold (\texttt{WordsBy.v})}
\label{fig:wordsby}
\end{figure}

@@

\begin{figure}
\begin{minted}{coq}
Definition MapThroughAlg{B : Set}(f:mappedT A B)
  : Alg (ListF A) (Const (list B)) :=
  rollAlg (fun R fo mapThrough xs => 
    match xs with
      Nil => []
    | Cons hd tl =>
      let (b,c) := f R fo hd tl in
        b :: mapThrough c
    end).
Definition mapThroughr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                      {B : Set}(f:mappedT A B) : R -> list B.
Definition mapThrough{B : Set}(f:mappedT A B) : List A -> list B.
\end{minted}
\caption{The algebra \texttt{MapThroughAlg} defining function \texttt{mapThrough} and \texttt{mapThroughr}; the
  code for those follows the pattern of \texttt{wordsBy} and \texttt{wordsByr} (Figure~\ref{fig:wordsby}), so
  we omit it (\texttt{MapThrough.v})}
\label{fig:mapthrough}
\end{figure}

@@

Assuming \verb|A : Set| and an equality test \verb|eqb : A -> A -> bool| on it,
we port this code to our Coq infrastructure in Figure~\ref{fig:rle}.
The function \verb|compressSpan| is written at the type \verb|mappedT A (nat * A)| that
will be required by \verb|mapThrough|.  Unfolding the definition of \verb|mappedT|,
\verb|compressSpan| has type:
\begin{minted}{coq}
forall(R : Set)(fo:FoldT (Alg (ListF A)) R), A -> R -> (nat * A) * R.
\end{minted}
\noindent It is invoked by the code for \verb|mapThrough| with \verb|fo : FoldT (Alg (ListF A)) R|.  It then has the
responsibility of extracting from the tail at type \verb|R| (second input) 
a result upon which \verb|mapThrough| should recurse (second component
of the output pair).  Then we define an algebra \verb|RleAlg| by
supplying \verb|compressSpan| as the function to map through, to
\verb|MapThroughAlg| (Figure~\ref{fig:mapthrough}).  Following the
pattern seen above, we define function \verb|rle| for top-level
recursions using \verb|fold| (we could also define a subsidiary version
\verb|rler|).

@@

The definition of \verb|Mu| in Figure~\ref{fig:mu} weakens
this to a strictly positive approximation. Instead of taking in \verb|F Mu|, constructor \verb|mu|
accepts an input of type \verb|F R|, for some type \verb|R| for which
we have a function of type \verb|R -> Mu|.  The impredicative
quantification of \verb|R| is essential here: we will instantiate it with
\verb|Mu| itself in the definition of \verb|inMu|
(Figure~\ref{fig:mu}).  So this approach would not work in a
predicative theory like Agda's.  The quantification of \verb|R| can be
seen as applying a technique due to Mendler, of introducing
universally quantified variables for problematic type occurrences, to
a datatype constructor.  We will review this in
Section~\ref{sec:related}.

@@

For a simple example: suppose we instantiate \verb|F| with
\verb|ListF| (of Figure~\ref{fig:listf}).  Our derivation uses a
different type that wraps \verb|F|, but this will show the issue in a
simple form.  Let us temporarily define \verb|List A| as
\verb|Mu (ListF A)| (again, for subsidiary recursion do not use just
\verb|ListF| directly).  The canonical way to define the empty list
would be, implicitly instantiating \verb|F| to \verb|ListF A|,
\begin{minted}{coq}
Definition mkNil := mu (List A) (fun x => x) (NilF A)
\end{minted}
\noindent But given this, there are infinitely many other equivalent
definitions.  For any \verb|Q : Set|, we could take
\begin{minted}{coq}
Definition mkNil' := mu Q (fun x => mkNil) (NilF A)
\end{minted}
\noindent Since \verb|fmap f (NilF A)| equals \verb|NilF B| for
\verb|f : A -> B|, if we apply \verb|outMu| (of Figure~\ref{fig:mu})
to \verb|mkNil'| or \verb|mkNil|, we will get \verb|NilF (List A)|.
But critically, \verb|mkNil| and \verb|mkNil'| are not equal, neither
definitionally nor provably.  One can define a function that puts
\verb|Mu| values in normal form by folding \verb|inMu| over them.
Then \verb|mkNil| and \verb|mkNil'| will have the same normal form,
and be equivalent in that sense.  But the fact that they are not
provably equal is what we term noncanonicity.

@@

Our work contributes to the program proposed by Owens and Slind, of
broadening the scope of functional programs that can be accommodated
in ITPs~\cite{owens+08}.  The goal of terminating recursion has been
advocated in the literature on programming languages under the name
\emph{strong functional programming}~\cite{turner95}.  Uustalu and
Vene developed a categorical view of a recursion scheme allowing one
level of subsidiary recursion, and illustrated it in Haskell with an
artificial example~\cite{uustalu11}.  In contrast, our scheme allows
arbitrary finite nestings of recursion, and we illustrate it in Coq
with realistic examples.  It seems that generalizing the carriers of
algebras to functors is the critical step enabling such examples. In
our Coq development, we show that Uustalu and Vene's scheme is
derivable from subsidiary recursion. Our method is also similar to the
technique of sized types, in providing a type-based method for
termination~\cite{barthe04}.  With sized types, datatypes are indexed
with abstract sizes, which must then be propagated through code, using
dependent types.  Our \verb|Subrec| does not require dependent types,
resulting in just polymorphically typed code for our examples
(\verb|Subreci|, for proofs, of course does use dependent types).

\textbf{Mendler-style recursion.}  Mendler introduced the idea of
using universal abstraction to support compositional termination
checking~\cite{mendler91}.  He proposed a functor-generic recursor of
type $\all{X}{(\all{R}{(R \to X) \to F\ R \to X}) \to \mu\ F\ \to
  X}$. We have applied this idea to the constructor of the type
\verb|Mu| (Section~\ref{sec:mu}).  Previous work explored the
categorical perspective on Mendler-style recursion, and showed how to
reduce it to basic catamorphisms (i.e., structural
recursion)~\cite{uustalu99}.  Another considered its use with negative
type schemes~\cite{ahn11}.  Previous work from our group showed how to
derive inductive datatypes in Cedille using extensions of the Mendler
encoding~\cite{firsov+18b,firsov+18a}.  Here, we do not derive
inductive types, but rather 
a terminating recursion scheme for existing datatypes.
